The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-20 paper/ms.tex
Add software citations
Add FI to ack
Small changes throughout

% \submitjournal{ApJ}

@@

\begin{figure*}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_one_orbit.pdf}
\end{center}
\caption{The artificial phase-dependence in the computed actions with an
error in the Galactocentric coordinate system. We consider here \thick{},
which has actions of $(J_R, J_{\phi}, J_z) = (37.9, -1520,
7.0)\,\actunit$ and $z_{\text{max}}=850\,\pc$ (see Table~\ref{tab:orbits}). We
integrate the orbit according to the procedure laid out in
Section~\ref{ssec:action_comp}, and which we plot in
Appendix~\ref{app:orbits}. Then, we subtract $100\,\pc$ from the $z$ value
(upper panels) or the $x$ value (lower panels) of each position in the orbit,
corresponding to an erroneous observer assuming a midplane (upper) or solar
radius (lower) that is off by $100\,\pc$. We then allow an observer
to measure the orbit over $1\,\Gyr$ and perform the same orbit integration
procedure at each timestep, and report the values of the actions, with the true values given as horizontal dashed lines. The
computation of $J_{\phi}$ is pristine to errors in $z$, with only numerical
artifacts remaining. Only small errors are induced in $J_R$, with the middle
$90\%$ of values over the $\Gyr$ being within $\sim8\%$ of the true $J_R$. As
expected, large errors are induced in $J_z$ with a $100\,\pc$ offset in $z$,
with the middle $90\%$ of values being within $\sim43\%$ of the true $J_z$.
The $x$ offset induces uncertainties in $J_R$, $J_{\phi}$, and $J_z$ of
$\sim21\%$, $\sim3\%$, and $\sim3\%$.}
\label{fig:one_orbit_wrong_ref}
\end{figure*}

@@

We consider the observer making a measurement, integrating an orbit, and
computing actions every 
    megayear 
using the prescription above. However, we
specify the star's starting position using the systematically offset
coordinate system.
Essentially we are shifting and then reintegrating at each point along the original orbit.
The actions computed using the offset coordinate system
for each phase-space starting point are shown for the first 
  gigayear 
of the orbit in the upper panels of
Figure~\ref{fig:one_orbit_wrong_ref}.\footnote{Occasionally the numerical
scheme fails and very large actions are reported by \texttt{gala}---we perform
a $5\sigma$ clip on each action to exclude such orbits, but this only excludes
a total of $5$ orbits out of the $1000$ considered for
Figure~\ref{fig:one_orbit_wrong_ref}. Some numerical artifacts remain, but the
vast majority of orbits are computed properly.}

@@

Figure~\ref{fig:midplane} shows the relative $z$ location of the inferred
midplane an observer would determine as a function of azimuth for
each galaxy, using their local solar neighborhood (the cylinder defined
above). The
$1\,\sigma$ error from the bootstrap procedure is shown as the dashed-line
error bars. The middle $90\%$ of midplane values across the solar circle spans
    $(1.9 \times 10^2, 1.6 \times 10^2, 84)\,\pc$ for these models.
In two of the three cases
the midplane therefore varies by more than $\pm 100\,\pc$ depending on the
azimuth along the solar circle; in the third (\mm{}, which has the thinnest
``thin disk'' of stars, but the largest stellar mass) the variation is closer
to~$\pm 50\,\pc$.

We compute the same midplane variation in Figure~\ref{fig:midplane_chervin},
but for four succcessive timesteps of the live N-body simulation of a
Sagittarius encounter \citep{2018MNRAS.481..286L}. Again we have subtracted a
best fit curve of the form 
    given in Equation~(\ref{eq:fit-curve}), with the values at times
    $(2.0, 4.0, 6.0, 6.9)\,\Gyr$ being $A = (9.5, 2.5, -2.1 \times 10^2,
    -3.9 \times 10^2)\,\pc$, $B
    = (0.074, 0.039, -36,
    -57)\,\text{deg}$, and $C = (2.6,-7.8, -65, -53)\,\pc$.
The middle $90\%$ of midplane values across the solar circle spans 
       $(49, 62, 140, 120)\,\pc$.

@@

\item 
    We show that such midplane variation is likely by measuring
the local galactic
midplane along the solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration, as well as a controlled simulation of the interaction of the Milky Way with Sagittarius. We found
that the midplane varies as a function of azimuth at the solar circle by
%mm [too much detail] $\sim185$, $162$, $84\,\pc$ (middle $90\%$) for
%the three cosmological simulations we considered (\mi{}, \mf{}, and
%\mm{}, respectively) and by $\sim60$ and $\sim120\,\pc$ before and
%after the interaction in the controlled simulation. 
   60--185~pc in these simulations.
%mm [this seems redundant with the next point.] Comparison of the corresponding velocity variations with recent measurements in the Milky Way suggests that these values are consistent with expectations for midplane variations in the Milky Way itself, which have not yet been measured as a function of azimuth.

@@

\acknowledgments
We would like to thank Megan Bedell, Robert A. Benjamin, Tobias Buck, Elena
D'Onghia, Benoit Famaey, and Adrian Price-Whelan for helpful discussions. A.B.
would like to thank Todd Phillips for helpful discussions. This project was
developed in part at the 2019 Santa Barbara Gaia Sprint, hosted by the Kavli
Institute for Theoretical Physics (KITP) at the University of California,
Santa Barbara. This research was supported in part at KITP by the
Heising-Simons Foundation and the National Science Foundation (NSF) under
Grant No. PHY-1748958. This work used the Extreme Science and Engineering
Discovery Environment (XSEDE), which is supported by NSF Grant No.
OCI-1053575. A.B. was supported in part by the Roy \& Diana Vagelos Program in
the Molecular Life Sciences and the Roy \& Diana Vagelos Challenge Award.
K.V.J.'s contributions were supported in part by the NSF under Grant No.
AST-1614743. M.-M.M.L. was partly supported by the NSF under Grant No.
AST-1815461.

>>>>

\submitjournal{ApJ}

@@

\begin{figure*}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_one_orbit.pdf}
\end{center}
\caption{The artificial phase-dependence in the computed actions with an
error in the Galactocentric coordinate system. We consider here \thick{},
which has actions of $(J_R, J_{\phi}, J_z) = (38, -1500,
7.0)\,\actunit$ and $z_{\text{max}}=850\,\pc$ (see Table~\ref{tab:orbits}). We
integrate the orbit according to the procedure laid out in
Section~\ref{ssec:action_comp}, and which we plot in
Appendix~\ref{app:orbits}. Then, we subtract $100\,\pc$ from the $z$ value
(upper panels) or the $x$ value (lower panels) of each position in the orbit,
corresponding to an erroneous observer assuming a midplane (upper) or solar
radius (lower) that is off by $100\,\pc$. We then allow an observer
to measure the orbit over $1\,\Gyr$ and perform the same orbit integration
procedure at each timestep, and report the values of the actions, with the true values given as horizontal dashed lines. The
computation of $J_{\phi}$ is pristine to errors in $z$, with only numerical
artifacts remaining. Only small errors are induced in $J_R$, with the middle
$90\%$ of values over the $\Gyr$ being within $\sim8\%$ of the true $J_R$. As
expected, large errors are induced in $J_z$ with a $100\,\pc$ offset in $z$,
with the middle $90\%$ of values being within $\sim43\%$ of the true $J_z$.
The $x$ offset induces uncertainties in $J_R$, $J_{\phi}$, and $J_z$ of
$\sim21\%$, $\sim3\%$, and $\sim3\%$.}
\label{fig:one_orbit_wrong_ref}
\end{figure*}

@@

We consider the observer making a measurement, integrating an orbit, and
computing actions every 
    megayear 
using the prescription above. However, we
specify the star's starting position using the systematically offset
coordinate system.
Essentially we are shifting and then reintegrating at each point along the original orbit.
The actions computed using the offset coordinate system
for each phase-space starting point are shown for the first 
  gigayear 
of the orbit in the upper panels of
Figure~\ref{fig:one_orbit_wrong_ref}.\footnote{Occasionally the numerical
scheme fails and very large actions are reported by \texttt{gala}---we perform
a $4\sigma$ clip on each action to exclude such orbits, but this only excludes
a total of $5$ orbits out of the $1000$ considered for
Figure~\ref{fig:one_orbit_wrong_ref}. Some numerical artifacts remain, but the
vast majority of orbits are computed properly.}

@@

Figure~\ref{fig:midplane} shows the relative $z$ location of the inferred
midplane an observer would determine as a function of azimuth for
each galaxy, using their local solar neighborhood (the cylinder defined
above). The
$1\,\sigma$ error from the bootstrap procedure is shown as the dashed-line
error bars. The middle $90\%$ of midplane values across the solar circle spans
    $(190, 160, 84)\,\pc$ for these models.
In two of the three cases
the midplane therefore varies by more than $\pm 100\,\pc$ depending on the
azimuth along the solar circle; in the third (\mm{}, which has the thinnest
``thin disk'' of stars, but the largest stellar mass) the variation is closer
to~$\pm 50\,\pc$.

We compute the same midplane variation in Figure~\ref{fig:midplane_chervin},
but for four succcessive timesteps of the live N-body simulation of a
Sagittarius encounter \citep{2018MNRAS.481..286L}. Again we have subtracted a
best fit curve of the form 
    given in Equation~(\ref{eq:fit-curve}), with the values at times
    $(2.0, 4.0, 6.0, 6.9)\,\Gyr$ being $A = (9.5, 2.5, -210,
    -390)\,\pc$, $B
    = (0.074, 0.039, -36,
    -57)\,\text{deg}$, and $C = (2.6,-7.8, -65, -53)\,\pc$.
The middle $90\%$ of midplane values across the solar circle spans 
       $(49, 62, 140, 120)\,\pc$.

@@

\item 
    We show that such midplane variation is likely by measuring
the local galactic
midplane along the solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration, as well as a controlled simulation of the interaction of the Milky Way with Sagittarius. We found
that the midplane varies as a function of azimuth at the solar circle by
   $60\textup{--}185\,\pc$ in these simulations.

@@

\acknowledgments
We would like to thank Megan Bedell, Robert A. Benjamin, Tobias Buck, Elena
D'Onghia, Benoit Famaey, and Adrian Price-Whelan for helpful discussions. A.B.
would like to thank Todd Phillips for helpful discussions. This project was
developed in part at the 2019 Santa Barbara Gaia Sprint, hosted by the Kavli
Institute for Theoretical Physics (KITP) at the University of California,
Santa Barbara. This research was supported in part at KITP by the
Heising-Simons Foundation and the National Science Foundation (NSF) under
Grant No. PHY-1748958. This work used the Extreme Science and Engineering
Discovery Environment (XSEDE), which is supported by NSF Grant No.
OCI-1053575. This work uses data hosted by the Flatiron Institute's FIRE data hub. A.B. was supported in part by the Roy \& Diana Vagelos Program in
the Molecular Life Sciences and the Roy \& Diana Vagelos Challenge Award.
K.V.J.'s contributions were supported in part by the NSF under Grant No.
AST-1614743. M.-M.M.L. was partly supported by the NSF under Grant No.
AST-1815461.

\software{This work made use of the following software: \texttt{astropy} \citep{astropy:2013,astropy:2018}, \texttt{gala}
\citep{gala, gala:zenodo}, \texttt{matplotlib} \citep{Hunter:2007}, \texttt{numpy} \citep{numpy:2011}, \texttt{scipy}
\citep{scipy:2001}, and \texttt{tqdm} \citep{casper_o_da_costa_luis_2019_2800317}.}
